---
title: "GroupB-NEW"
author: "Samara Tanz"
date: "2024-11-20"
output: pdf_document
---
# Loading packages
```{r}
library(vegan)
library(tidyverse)
library(deSolve)
library(fitdistrplus)
library(flexsurv)
library(lme4)
library(car)
library(rgl)
library(ggfortify)

options(rgl.useNULL = TRUE) 
rgl::setupKnitr(autoprint = TRUE)
```

# Loading csv files
```{r}
#Provided original csv files (separated sheets from the master excel file)
Plant_traits <- read.csv("Plant_traits.csv")
Arthropod_List <- read.csv("Arthropod_List.csv")
Stable_Isotope <- read.csv("Stable_Isotope.csv")
Soil_traits <- read.csv("Soil_traits.csv")
```

## Creating New Arthropod list
```{r}
#Updated arthropod list which switched the rows and columns of treatment and 
# species while neglecting every other irrelevant column 

new_insect <- Arthropod_List  %>% 
  gather(var, val, 2:ncol(Arthropod_List)) %>% 
  spread(Species, val)
  

Arthro_List_update <-Arthropod_List [-c(1,2,3), ]
```
```{r}
write.csv(Arthro_List_update, 'New_insect.csv')

test_data <- read.csv("final_arthro.csv")
```


```{r}
Arthro_List_update <- read.csv("final_arthro.csv") 
```


# Shannon Diversity Calculations
```{r}
#Excluding the first and last column (as they d not contain data)
fixed_arthro = Arthro_List_update[,2:350] 

#Shannon diversity 
shannon_indicies <- diversity(fixed_arthro, index = "shannon")

#Reformatting with column names specified 
final_shannon1 <- data.frame(Treatment=Arthro_List_update$Plot,
                            Shannon=shannon_indicies)
```

# Histogram view 
Preliminary to determining distribution type.
```{r}
# With added density curve
hist(final_shannon1$Shannon, probability = TRUE, col = "lightblue", border = "black", breaks = 25)
lines(density(final_shannon1$Shannon), col = "red", lwd = 2)  # Adds a density curve
```
This is normal looking, however we need to determine distribution type in order to make the correct analysis and conclusions about our findings...

#Organizing
##Group Treatments into Communities
```{r}
#Data is separated by into 75 treatments -> groups with 15 treatments each
#To draw better conclusions, we need to group by group
final_shannon1$Community <- ifelse(startsWith(final_shannon1$Treatment, "OP"), "OP",
                                          ifelse(startsWith(final_shannon1$Treatment, "RP"), "RP",
                                                 ifelse(startsWith(final_shannon1$Treatment, "TP"), "TP",
                                                        ifelse(startsWith(final_shannon1$Treatment, "IS"), "IS",
                                                               ifelse(startsWith(final_shannon1$Treatment, "PS"), "PS", "Other")))))

#Rearranging columns so "Community" is beside "Treatment"
shannon_reg <- final_shannon1[, c("Treatment", "Community", "Shannon")]
```

```{r}
#Visualization of Shannon diversity for each community 
ggplot(shannon_reg, aes(x = Community, y = Shannon)) +
  geom_bar(stat = "identity", fill = "yellowgreen", color = "black") +
  theme_minimal() +
  labs(title = "Shannon Diversity for each Community", x = "Community", y = "Shannon Diversity")
```
#Add all the predictor variable columns (soil and plant traits) onto final_shannon...
```{r}
final_shannon1 = data.frame(Treatment = Arthro_List_update$Plot,
                           Shannon = shannon_indicies)
```

..... HELP

```{r}
#Save dataframe as a csv file for sharing and future edits...
write.csv(final_shannon1, "final_shannon.csv")
```

#Filtering data for our final dataset
```{r}
#OP14 and OP15 have N/A for leafN (their had a technical error) 
noNA_FS <- filter(final_shannon, !is.na(final_shannon$leaf_N))
```
```{r}
#Replacing those cells with the mean of the total leafN 
data <- final_shannon 

data[data$Treatment == "OP14", "leaf_N"] <- 
  round(mean(noNA_FS$leaf_N), 2)

data[data$Treatment == "OP15", "leaf_N"] <- 
  round(mean(noNA_FS$leaf_N), 2)
```

#Determining distribution type
We test for normal, gamma, and inverse gaussian becauseADD SOMETHING HERE 
```{r}
#Normal
# MLE: LogLik = -19.28128, mean = 3.43, sd = 0.31

norm_combos <- expand.grid(mean = seq(2.4, 4.2, 0.01),
                            sd = seq(0.15, 0.7, 0.01))
```

```{r}
#Gamma
# MLE: LogLik = -21.03414, shape = 98, scale = 0.035

gamma_combos <- expand.grid(shape = seq(98, 100, 0.001),
                            scale = seq(0.03, 0.04, 0.001))

```

```{r}
#Inverse Gaussian 
# MLE: LogLik = -21.50412, mu =	3.43, lambda = 381.81

invgaus_combos <- expand.grid(mu = seq(2.4, 4.2, 0.01),
                              lambda = seq(375, 385, 0.01))
	
```

## Log-Likelihood evaluators
```{r}
norm_loglik <- function(x = data$Shannon, mean, sd){
  return (sum(log(dnorm(x = x, mean = mean, sd = sd))))
}


gamma_loglik <- function(x = data$Shannon, shape, scale){
  return (sum(log(dgamma(x = x, shape = shape, scale = scale))))
}


invgaus_loglik <- function(x = data$Shannon, mu, lambda){
  return (sum(log(
    sqrt(lambda/(2*pi*x^3)) * exp((-lambda*(x-mu)^2)/(2*x*mu^2))
    )))
}
```

```{r}

LogLik_norm <- c() # empty log-likelihoods vector for NORMAL

for (i in 1:nrow(norm_combos)){
  LogLik_norm[i] <- norm_loglik(mean = norm_combos$mean[i],
                                sd = norm_combos$sd[i])
}

LLs_norm <- data.frame(LogLik_Norm = LogLik_norm,
                       mean = norm_combos$mean,
                       sd = norm_combos$sd)

LLs_norm |> subset(LogLik_Norm == max(LogLik_Norm)) -> MLENorm
```

```{r}

LogLik_gamma <- c() # empty log-likelihoods vector for GAMMA

for (i in 1:nrow(gamma_combos)){
  LogLik_gamma[i] <- gamma_loglik(shape = gamma_combos$shape[i],
                                  scale = gamma_combos$scale[i])
}

LLs_gamma <- data.frame(LogLik_Gamma = LogLik_gamma,
                       shape = gamma_combos$shape,
                       scale = gamma_combos$scale)

LLs_gamma |> subset(LogLik_gamma == max(LogLik_gamma)) -> MLEGamma
```

```{r}

LogLik_invgaus <- c() # empty log-likelihoods vector for INVERSE GAUSSIAN

for (i in 1:nrow(invgaus_combos)){
  LogLik_invgaus[i] <- invgaus_loglik(mu = invgaus_combos$mu[i],
                                      lambda = invgaus_combos$lambda[i])
}

LLs_invgaus <- data.frame(LogLik_InvGaus = LogLik_invgaus,
                          mu = invgaus_combos$mu,
                          lambda = invgaus_combos$lambda)

LLs_invgaus |> subset(LogLik_invgaus == max(LogLik_invgaus)) -> MLEInvgaus
```


```{r}
LLs_norm |> subset(sd == MLENorm$sd) |> 
  subset(abs(LogLik_Norm - MLENorm$LogLik_Norm) < 1.92) |>
  summarise(lowerCI_mean = min(mean), upperCI_mean = max(mean)) -> CI_mean

LLs_norm |> subset(mean == MLENorm$mean) |> 
  subset(abs(LogLik_Norm - MLENorm$LogLik_Norm) < 1.92) |>
  summarise(lowerCI_sd = min(sd), upperCI_sd = max(sd)) -> CI_sd

LLs_norm |> ggplot(aes(x = mean, y = sd, z = LogLik_Norm, fill = LogLik_Norm)) + 
  geom_raster() + geom_contour(color = "grey") +
  geom_pointrange(inherit.aes = F, aes(x = MLENorm$mean, 
                                       y = MLENorm$sd,
                                       ymin = CI_sd$lowerCI_sd,
                                       ymax = CI_sd$upperCI_sd), color = "red") +
  geom_pointrange(inherit.aes = F, aes(x = MLENorm$mean, 
                                       xmin = CI_mean$lowerCI_mean,
                                       xmax = CI_mean$upperCI_mean,
                                       y = MLENorm$sd), color = "red")

LLs_norm |> subset(sd == MLENorm$sd) |> ggplot(aes(x = mean, y = LogLik_Norm)) + geom_line()
LLs_norm |> subset(mean == MLENorm$mean) |> ggplot(aes(x = sd, y = LogLik_Norm)) + geom_line()
```

```{r}
LLs_gamma |> subset(shape == MLEGamma$shape) |> 
  subset(abs(LogLik_Gamma - MLEGamma$LogLik_Gamma) < 1.92) |>
  summarise(lowerCI_scale = min(scale), upperCI_scale = max(scale)) -> CI_scale

LLs_gamma |> subset(scale == MLEGamma$scale) |> 
  subset(abs(LogLik_Gamma - MLEGamma$LogLik_Gamma) < 1.92) |>
  summarise(lowerCI_shape = min(shape), upperCI_shape = max(shape)) -> CI_shape

LLs_gamma |> ggplot(aes(x = scale, y = shape, z = LogLik_Gamma, fill = LogLik_Gamma)) + 
  geom_raster() + geom_contour(color = "grey") +
  geom_pointrange(inherit.aes = F, aes(x = MLEGamma$scale, 
                                       y = MLEGamma$shape,
                                       ymin = CI_shape$lowerCI_shape,
                                       ymax = CI_shape$upperCI_shape), color = "red") +
  geom_pointrange(inherit.aes = F, aes(x = MLEGamma$scale, 
                                       xmin = CI_scale$lowerCI_scale,
                                       xmax = CI_scale$upperCI_scale,
                                       y = MLEGamma$shape), color = "red")

LLs_gamma |> subset(shape == MLEGamma$shape) |> ggplot(aes(x = scale, y = LogLik_Gamma)) + geom_line()
LLs_gamma |> subset(scale == MLEGamma$scale) |> ggplot(aes(x = shape, y = LogLik_Gamma)) + geom_line()
```

```{r}
LLs_invgaus |> subset(mu == MLEInvgaus$mu) |> 
  subset(abs(LogLik_InvGaus - MLEInvgaus$LogLik_InvGaus) < 1.92) |>
  summarise(lowerCI_lambda = min(lambda), upperCI_lambda = max(lambda)) -> CI_lambda

LLs_invgaus |> subset(lambda == MLEInvgaus$lambda) |> 
  subset(abs(LogLik_InvGaus - MLEInvgaus$LogLik_InvGaus) < 1.92) |>
  summarise(lowerCI_mu = min(mu), upperCI_mu = max(mu)) -> CI_mu

LLs_invgaus |> ggplot(aes(x = mu, y = lambda, z = LogLik_InvGaus, fill = LogLik_InvGaus)) + 
  geom_raster() + geom_contour(color = "grey") +
  geom_pointrange(inherit.aes = F, aes(x = MLEInvgaus$mu, 
                                       y = MLEInvgaus$lambda,
                                       ymin = CI_lambda$lowerCI_lambda,
                                       ymax = CI_lambda$upperCI_lambda), color = "red") +
  geom_pointrange(inherit.aes = F, aes(x = MLEInvgaus$mu, 
                                       xmin = CI_mu$lowerCI_mu,
                                       xmax = CI_mu$upperCI_mu,
                                       y = MLEInvgaus$lambda), color = "red")

LLs_invgaus |> subset(mu == MLEInvgaus$mu) |> ggplot(aes(x = lambda, y = LogLik_InvGaus)) + geom_line()
LLs_invgaus |> subset(lambda == MLEInvgaus$lambda) |> ggplot(aes(x = mu, y = LogLik_InvGaus)) + geom_line()
```

##AIC
```{r}
-2*MLENorm$LogLik_Norm + 2*2
-2*MLEGamma$LogLik_Gamma + 2*2
-2*MLEInvgaus$LogLik_InvGaus + 2*2

dinvgaus <- function(x, mu, lambda){
  return (
    sqrt(lambda/(2*pi*x^3)) * exp((-lambda*(x-mu)^2)/(2*x*mu^2))
    )
}

vals <- seq(min(data$Shannon),max(data$Shannon), length = 1000)

NormalDen <- cbind(x = vals, prob = dnorm(vals, mean = MLENorm$mean, sd = MLENorm$sd))
GammaDen <- cbind(x = vals, prob = dgamma(vals, shape = MLEGamma$shape, scale = MLEGamma$scale))
InverseGammaDen <- cbind(x = vals, 
                         prob = dinvgaus(vals, mu = MLEInvgaus$mu, lambda = MLEInvgaus$lambda))

data |> ggplot(aes(x = Shannon)) + geom_rug() +
  geom_line(data = NormalDen, aes(x = x, y = prob), color = "red") +
  geom_line(data = GammaDen, aes(x = x, y = prob), color = "blue") +
  geom_line(data = InverseGammaDen, aes(x = x, y = prob), color = "black")
```


# Find which varaiables are linearly and positively correlated... 
```{r}
par(mfrow=c(1,1))

pairs(data[, c("soil_C", "soil_N", "soil_P", "soil_pH", "soil_sal", "soil_wat", 
               "biomass_above", "plant_dens", "leaf_N", "leaf_C", "leaf_P")])
```

#Test for co-linearality?
```{r}
cor_matrix <- cor(data[, c("soil_C", "soil_N", "soil_P", "soil_pH", "soil_sal",
                           "soil_wat", "biomass_above", "plant_dens", "leaf_N", 
                           "leaf_C", "leaf_P")], 
                  use = "pairwise.complete.obs")
print(cor_matrix)
```

#PCA!
```{r}
#PCA of all traits 
pca.out <- prcomp(data[, c("soil_C", "soil_N", "soil_P", "soil_pH", "soil_sal", 
                           "soil_wat", "biomass_above", "plant_dens", "leaf_N", 
                           "leaf_C", "leaf_P")], scale.=T)
```

#This is not working for me :(
```{r}
#as.data.frame(pca.out$var$contrib)
```

# PC1: (1) soil_N (-0.3543157) (2) soil_wat (-0.3536549) (3) soil_pH (0.3370312) (4) biomass_above (0.3269548) (5) soil_C (-0.3267151)
# PC2: (1) soil_C (0.44891398) (2) plant_dens (-0.40295140) (3) soil_sal (0.39164170) (4) soil_N (0.37709659)
```{r}
pca.out
```

```{r}
summary(pca.out)
```

```{r}
head(pca.out$rotation)
```

```{r}
#Visual of PCA
autoplot(pca.out, x=1,y=2,
         data=final_shannon_UPDATED, 
         colour="Treatment_grouped",
         frame=TRUE) +
  theme_classic()
```

