---
title: "Group Project Assignment"
author: "Samara Tanz"
date: "2024-09-26"
output: pdf_document
---

```{r}
install.packages("vegan")
library(vegan)
library(tidyverse)
library(deSolve)
install.packages("fitdistrplus")
library(fitdistrplus)
install.packages("flexsurv")
library(flexsurv)
```

```{r}
Plant_traits <- read.csv("Plant_traits.csv")
Arthropod_List <- read.csv("Arthropod_List.csv")
Stable_Isotope <- read.csv("Stable_Isotope.csv")
Soil_traits <- read.csv("Soil_traits.csv")
```
1. Calculating diversity --> Shannon diversity 
--> using csv file Gaby made that switched the rows and columns of the arthropod list 
```{r}
#Manual reformatted version of the Arthropod List data for one species per 
# column along with one group per row as the "Treatment"
Species <- read.csv("final_arthro.csv") 
```
Jared and Gaby's reformat of group number and Species (1 species per column specified by each row)

```{r}
#Excluding the first and last column 
species_data = Species[,2:350] 

#Using the 'diversity()' function to get Shannon diversity 
shannon_diversity <- diversity(species_data, index = "shannon")

# Reformatting with column names specified 
final_shannon <- data.frame(Treatment=Species$Plot,
                            Shannon=shannon_diversity)
```
as the first column contains the group name and the last column is species "x" with no data included

View the results --> this is w/o column/row names 
view(shannon_diversity) #YES

view(final_shannon)

2. Model selection to figure out which to use for our statistics --> is the data normal?
```{r}
#Histogram view
final_shannon |> ggplot(aes(x=Treatment, y=Shannon)) + 
  geom_bar(stat="identity", fill="skyblue", colour="black")
```
```{r}
#Using the Shapiro test to test normality with p-value
shapiro.test(shannon_diversity) 
```
the data is not normal bc p-value=0.06938 which is >0.05
... which model is best to use then...?~~>
uh oh... not normally distributed/significant --> _general linear mixed model_ w AICc scores 
--> logistic is the best based on the AICc values Jared found:

```{r}
# Fit a logistic distribution to the Shannon diversity data
logisticfit <- fitdist(final_shannon$Shannon, "logis")

# Generate a QQ plot for the logistic distribution fit
qqcomp(list(logisticfit), legendtext = "logistic")
```
```{r}
# fitting different distributions to that Shannon diversity 
gammafit <- fitdist(final_shannon$Shannon, "gamma")
weibullfit <- fitdist(final_shannon$Shannon, "weibull")
logisticfit <- fitdist(final_shannon$Shannon, "logis")
lnormfit <- fitdist(final_shannon$Shannon, "lnorm")
gengammafit <- fitdist(final_shannon$Shannon, "gengamma", 
                       start = list(mu = mean(final_shannon$Shannon), 
                                    sigma = sd(final_shannon$Shannon), Q = 0))

```
```{r}
# Extract AIC and BIC for each distribution
aic_bic_table <- data.frame(
  Distribution = c("Gamma", "Weibull", "Logistic", "Lognormal", "Generalized Gamma"),
  AIC = c(gammafit$aic, weibullfit$aic, logisticfit$aic, lnormfit$aic, gengammafit$aic),
  BIC = c(gammafit$bic, weibullfit$bic, logisticfit$bic, lnormfit$bic, gengammafit$bic))
```

```{r}
# Display AIC and BIC values for comparison
print(aic_bic_table)
```

general mixed vs mixed --> general means not linear, mixed means random and not mixed 
--> general means it can be normal/non-normal (rather than w mixed model it has to be normal)
--> recall the beta stuff (beta0 + beta1 + beta2), each beta is a coefficient for each variable we have
  -> i.g 'plant traits' and 'soil traits' consist of a lot of variables, we want to find the best coefficient to represent the data (recall: lmr)


the data being logistic (bc of the test we ran on it) makes us use _"glmr"_

3. Create a general LMM (linear mixed model)

what we're gonna be asking ~> are plant or soil traits the cause of the differences we see?
--> we'll test this out by measuring each of the traits in plant and soil documents 

4. Interpret to see which plant/soil traits/factors are most significant 
(2 models: soil and plants)

